Robot Team III: Scotty Strikes Back

### 2024-04-28 paper/outline.tex
updated with equations
pushing changes to overleaf

NOTE: We should move the description of the GTM algroithm to the top of the methods section.

\section*{Outline}
\begin{outline}[enumerate]
\1 Introduction
    \2 Characterization of inland water quality via remote sensing is challenging for 4 key reasons:
        \3 Key limitations:
            \4 Spectral resolution (visible and multi-spectral imaging)
            \4 Spatial Resolution (large pixel sizes)
            \4 Temporal Resolution (long satellite revisit times)
            \4 Availability of in situ reference data
        \3 When appropriate data are available, supervised ML approaches can map reflectance from remote sensing imagery to relevant water quality parameters
    \2 Hyperspectral Imaging addresses spectral resolution
    \2 UAV adresses spatial and temporal resolution
    \2 Robot team addresses in situ data... but what do we do if we don't already know which sensors to use? 
    \2 Unsupervised ML, Endmember extraction, and Spectral Unmixing
        \3 k-NN, fuzzy c-means, matrix factorizations, etc.
            \4 Are general but do not provide relationship between classes e.g. is class 1 more similar to class 2 or class 10?
        \3 SOM addresses this by enforcing a topological relationship between clusters. Classes close to each other in the map are more similar than classes far apart
        \3 Similarly, we may also be interested in the spectral characteristics of the learned *classes*, i.e. we want to think of the classes as *endmembers*.
            \4 Traditional approaches rely on assumption of existence of "pure pixels" in the scene
            \4 additionally, we may have to treat number of endmembers as a hyperparameter

>>>>

NOTE: We should move the description of the GTM algorithm to the top of the methods section.

\section*{Outline}
We should recast the introduction to focus on unsupervised classification rather than endmember extraction...

Characterization on inland waters via remote sensing is hard because (reasons). Hyperspectral imaging has emerged as a pivotal technology in this space due to the dramatic improvement in spectral resolution afforded by HSI platforms. Specifically, the combination with hyperspectral imaging together with autonomous collection by UAVs enables unparalleled spectral, spatial, and temporal resolution. Significant effort has developed the use of hyperspectral imagery for the inversion of optically-active water quality parameters. However, this approach requires access to sufficient quantities of in situ reference data to fit and evaluate each model. Traditionally, this process has required collocation of remote sensing imagery with data from field campaigns spanning decades (cite Aurin and Ross). Recently, we have demonstrated how combining autonomous collection of in situ data with UAV-borne hyperspectral imagery dramatically reduces the time required to collect these data. Both of these approaches require we already know what to look for.

\begin{outline}[enumerate]
\1 Introduction
    \2 Characterization of inland water quality via remote sensing is challenging for 4 key reasons:
        \3 Key limitations:
            \4 Spectral resolution (visible and multi-spectral imaging)
            \4 Spatial Resolution (large pixel sizes)
            \4 Temporal Resolution (long satellite revisit times)
            \4 Availability of in situ reference data
        \3 When appropriate data are available, supervised ML approaches can map reflectance from remote sensing imagery to relevant water quality parameters
    \2 Hyperspectral Imaging addresses spectral resolution
    \2 UAV addresses spatial and temporal resolution
    \2 Robot team addresses in situ data... but what do we do if we don't already know which sensors to use? 
    \2 Unsupervised ML, Endmember extraction, and Spectral Unmixing
        \3 k-NN, fuzzy c-means, matrix factorizations, etc.
            \4 Are general but do not provide relationship between classes e.g. is class 1 more similar to class 2 or class 10?
        \3 SOM addresses this by enforcing a topological relationship between clusters. Classes close to each other in the map are more similar than classes far apart
        \3 Similarly, we may also be interested in the spectral characteristics of the learned *classes*, i.e. we want to think of the classes as *endmembers*.
            \4 Traditional approaches rely on assumption of existence of "pure pixels" in the scene
            \4 additionally, we may have to treat number of endmembers as a hyperparameter

### 2024-04-28 paper/paper.tex
pushing changes to overleaf

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Unmanned Aerial Vehicles (UAVs) equipped with hyperspectral imagers have emerged as an essential technology for the characterization of inland water bodies. The high spectral and spatial resolutions of these systems enable the retrieval of a plethora of optically-active water quality parameters via band ratio algorithms and machine learning methods. However, fitting and validating these models requires access to sufficient quantities of in situ reference data which are time-consuming and expensive to obtain. In this study, we demonstrate how the Generative Topographic Mapping (GTM), a Bayesian realisation of the Self Organizing Map (SOM), enables highly-detailed unsupervised classification and nonlinear endmember extraction of UAV acquired hyperspectral imagery. Specifically, the topographic relationship between classes maintained by the GTM presents an attractive alternative to other unsupervised approaches which do not provide a notion of similarity between learned classes. Using data collected across a North Texas pond, we apply the GTM to create a detailed map segmenting the area into distinct regions of interest. Further, by interpreting the learned classes as distinct endmembers, the GTM can be used to rapidly identify and map unique water constituents. We demonstrate this capability by using the GTM to map algal abundance as well as the dispersion of a rhodamine dye tracer released into the water.
}

% Keywords
\keyword{Hyperspectral Imaging; Remote Sensing; Unsupervised Classification; Endmember Extraction, Spectral Unmixing}

@@

@@

\noindent 
\begin{tabular}{@{}ll}
UAV & Unmanned Aerial Vehicle \\
SOM & Self Organizing Map \\ 
GTM & Generative Topographic Mapping  \\
HSI & Hyperspectral Image \\ 
CDOM & Colored Dissolved Organic Matter \\
HAB & Harmful Algal Bloom 
\end{tabular}
}

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{FIXME}

% Keywords
\keyword{Hyperspectral Imaging; Remote Sensing; Unsupervised Classification; Spectral Unmixing}

@@

Inland water bodies pose a unique challenge for analysis by remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability; nevertheless, significant effort has gone into the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, turbidity, and temperature (citation). Machine learning techniques have proven to be a powerful tool in this domain, facilitating the development of data-driven models which map observed reflectance to parameters of interest. However, these methods rely on the availability of significant volumes of in-situ reference data needed to calibrate models and asses their uncertainty. For example Aurin et al... Similarly, Ross et al developed... The challenge posed by in-situ data collection together with long satellite revisit times limits the utility of remote sensing imagery for real-time applications.

- supervised classification of water quality with XGBoost \cite{ghatkar2019classification}
- Spectral signatures of classes are complex and often overlapping \cite{thenkabail2018hyperspectral}
- Remote sensing used for oil spill analysis: extent and thickness mapping \cite{kokaly2013spectroscopic, leifer2012state}
- Sun glitter remains a key challenge for sensing in the visible portion of the spectrum but multi and hyperspectral imagers have been used for oil spill identification and to identify their impacts on vegetation stress and mortality \cite{fingas2014review, khan2018modern}
- common approach is inversion of optically-active water quality parameters such as chlorophyll-a, blue-green algae, turbidity, and temperature \cite{ritchie2003remote}
- Combining spectral indices such as the NDVI together with machine learning is a popular approach \cite{thenkabail2018hyperspectral, sagan2020monitoring, lu2021retrieval}
- Polynomial regression models for chlorophyll-a, turbidity \cite{zhang2022selection}
- Ross et al. created a comprehensive dataset with over 600,000 water quality records matching optically active water quality parameters with associated satellite imagery from Landsat 5,7,8. To achieve this quantity of data, they needed records spanning 1984 to 2019 \cite{ross2019aquasat}
- Aurin et al. took a similar route combining 30 years of remote sensing imagery with in-situ data from over 500 field compaigns for CDOM, organic carbon, etc... \cite{aurin2018remote}.
- When ground truth data aren't available in sufficient quantity, classification into water quality categories is another approach (ground truth can be easier to obtain by expert analysis of scene) \cite{koponen2002lake}

Unsupervised classification
- When no ground-truth data are available, unsupervised classification can still help partition imagery into groups or clusters.
- Many data-driven, ML methods have been employed for the task including various matrix factorizations, k-nearest neighbors, fuzzy c-means, density estimation methods, etc. \cite{zhang2019hyperspectral}
- Additionally unsupervised approaches can be used to perform nonlinear dimensionality reduction & pre-processing for supervised approaches.
- Many unsupervised approaches can be challenging to interpret due to the lack of a relationship between the identified classes, e.g. is class 1 more similar to class 2 than class 10?
- The Self Organizing Map is a popular technique which classifies data while enforcing topological relationships between the resulting classes such that classes close to each  other in the latent space are more similar than classes that are far apart.
- SOM used for remote sensing imagery classification \cite{msom-remote-sensing}
- SOM used for clustering and data compression of HSI cube-sat \cite{danielsen2021self}
- SOM used for land-use and land-cover change analysis \cite{penfound2021analysis}
- SOM has bee nused for identifying synoptic-scale patterns in wind and sea surface temperature data {som-satellite}

endmember extraction, spectral unmixing
- This is a related problem where the goal is to identify unique spectral signatures which combine (linearly or non-linearly) to produce the measured signal.
- Having identified the set of endmembers, we then seek to determine their relative abundance in each pixel
- Many statistical approaches for extracting endmembers \cite{berman2004ice}
- Sparse PCA used for endmember extraction \cite{yousefi2016mineral}
- Popular spectral unmixing approach is to treat "pure" endmembers as vertices of a simplex \cite{plaza2012endmember, nascimento2005vertex}
- Convolutional Neural Networks with Autoencoder architectures are a popular ML approach which identify endmembers and perform non-linear unmixing \cite{palsson2020convolutional, non-negative-autoencoders, su2019daen, borsoi2019deep}
- Self Organizing Map is another approach which has been used for endmember extraction together with neural networks for abundance mapping (unmixing) \cite{cantero2004analysis}

Hyperspectral imaging has emerged as a pivotal technology in this space. With their increased spectral resolution, hyperspectral imagers (HSIs) capture the complex chemical, biological, environmental characteristics of reflectance spectra. (add references for specific applications of HSIs). Continued development of HSI technology has led to a dramatic reduction in their size enabling their utilize as the payload of small autonomous aerial vehicles (UAVs) (citation). By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simulatenously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).

- applications include food quality & safety, medical diagnoses, precision agriculture, and forensic document examination \cite{khan2018modern}
- Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}
- Near-earth HSI addresses spatial, spectral, and temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping \cite{banerjee2020uav}
- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
- UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU \cite{arroyo2019implementation}
- UAV-based HSI for turbidity estimation \cite{vogt2016near}

In our previous work, we introduced a robot team designed to rapidly characterize water composition in novel environments \cite{robot-team-1}. 
Joining UAV-borne HSI together with collection of in-situ reference data by an autonomous boat dramatically reduces the time required to generate .

By carefully coordinating the simultanous collection of ground-truth water quality data and hyperspectral imagery, this team can rapidly estimate concentrations of optically-active components including colored dissolved organic matter (CDOM), blue-green algae, crude oil, temperature as well

In this paper, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian alternative to the Self Organizing Map, can be utilized to rapidly identify relevant spectral signatures within a scene. Using data collected at site in Montague, North Texas, we demonstrate the ability of this paradigm with two case studies. First, we demonstrate how the GTM can be used to provide a high-resolution unsupervised classification of a water body enabling the rapid identification of relevant regions which can then be used to direct further investigation by reference-sensor equipped boats. Second, we demonstrate the ability to perform endmember extraction and abundance mapping using imagery from a test-release of rhodamine dye. Together with the normalized spectral similarity score, these spectral signatures are identified and their associated area in the scene is determined.

@@

\noindent 
\begin{tabular}{@{}ll}
HAB & Harmful Algal Bloom \\ 
HSI & Hyperspectral Image \\ 
SOM & Self Organizing Map \\ 
GTM & Generative Topographic Mapping 
\end{tabular}
}
